Exploring Anomalies in Pacific Northwest Special Education Trends

### 2020-12-06 scripts/Final_project.Rmd
Deleted splom

```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
library("papaja")
r_refs("r-references.bib")
library(here)
library(tidyverse)
library("readxl")
library(janitor)
library(esc)
library(dplyr)
install.packages("splom")

>>>>

```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
library("papaja")
r_refs("r-references.bib")
library(here)
library(tidyverse)
library("readxl")
library(janitor)
library(esc)
library(dplyr)

### 2020-12-06 scripts/Final_project.tex
Deleted splom

---
title             : "Exploring Anomalies in Pacific Northwest Special Education Trends"
shorttitle        : "PACNW SpEd Trends"

author: 
  - name          : "David Fainstein"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "1655 Alder, Eugene, Oregon, USA  97403"
    email         : "dfainste@uoregon.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Janette Avelar"
    affiliation   : "1"
    role:
      - Writing - Review & Editing
  - name          : "Makayla Whitney"
    affiliation   : "1"
    role:
      - Writing - Review & Editing
  - name          : "Joe Swinehart"
    affiliation   : "1"
    role:
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of Oregon"

authornote: |
  Janette Avelar, University of Oregon College of Education

@@

  Joe Swinehart, Behavioral Research & Teaching

@@

  Contact: dfainste@uoregon.edu
abstract: |
  This paper will examine the changes in public school special education populations in Vancouver, British Columbia and Oregon over the last 20 years. The British Columbia (BC) data was pulled from an open-source data catalogue, published by the education analytics division of the Ministry of Education in British Columbia. This longitudinal data acquistion is part of formal student status reporting that is required within BC public schools - typically occuring in October of each school year. The Oregon data was pulled from the Oregon Department of Education and covers special education enrollment from 2002 - 2020. This data acquistion is part of formal student status reporting that is required for compliance with federal IDEA laws - typically occuring in November or February of each school year. Within the BC dataset, students with exceptionalities are categorized in one of 12 eligibility categories; within the Oregon dataset, this is 13 categories.

keywords          : "Vancouver, British Columbia, Oregon, Special Education, SpEd"
wordcount         : "0<n<1,000,000"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
library("papaja")
r_refs("r-references.bib")
library(here)
library(tidyverse)
library("readxl")
library(janitor)
library(esc)
library(dplyr)

BCdata <- read_excel(here("data", "bcsped_headcount.xlsx"))

BCpopulation <- read_excel(here("data", "BCpopdata.xlsx")) %>% 
  select("SCHOOL_YEAR",
         "TOTAL_ENROLMENT",
         "PUBLIC_OR_INDEPENDENT") %>% 
  filter(PUBLIC_OR_INDEPENDENT == "BC PUBLIC SCHOOL") %>% 
  mutate(SCHOOL_YEAR = as.factor(SCHOOL_YEAR), 
         SCHOOL_YEAR = as.numeric(SCHOOL_YEAR)+8)

BCpop_district <- read_excel(here("data", "BCdistrictpop.xlsx")) %>% 
  select("SCHOOL_YEAR",
         "TOTAL_ENROLMENT",
         "DISTRICT_NUMBER") %>% 
  mutate(SCHOOL_YEAR = as.factor(SCHOOL_YEAR), 
         SCHOOL_YEAR = as.numeric(SCHOOL_YEAR)+8)

BCdatastripped <- BCdata %>% 
    #Makayla & Janette - may want to include more columns here with the select for our global "BCdatastripped" df
    select("SCHOOL_YEAR", 
         "DATA_LEVEL", 
         "SPECIAL_NEEDS_CODE", 
         "NUMBER_OF_STUDENTS", 
         "PUBLIC_OR_INDEPENDENT",
         "DISTRICT_NUMBER") %>% 
  filter(SPECIAL_NEEDS_CODE != "M",
         SPECIAL_NEEDS_CODE != "J",
         SPECIAL_NEEDS_CODE != "N") %>% 
    mutate(SPECIAL_NEEDS_CODE = fct_recode(SPECIAL_NEEDS_CODE,
                          "Physically Dependent" = "A",
                          "Deafblind" = "B",
                          "Moderate+ ID" = "C",
                          "Chronic Health" = "D",
                          "Visual Impairment" = "E",
                          "Deaf or Hard of Hearing" = "F",
                          "Autism" = "G",
                          "High Behavioural Need" = "H",
                          "Mild ID" = "K",
                          "Learning Disability" = "Q",
                          "Gifted" = "P",
                          "Mild Behavioural Need" = "R"),
           SCHOOL_YEAR = as.factor(SCHOOL_YEAR), 
           NUMBER_OF_STUDENTS = as.integer(NUMBER_OF_STUDENTS),
           SCHOOL_YEAR = as.numeric(SCHOOL_YEAR)) %>% 
  filter(SCHOOL_YEAR >= 9)

BC_districtdata <- BCdatastripped %>% 
  filter(DATA_LEVEL == "DISTRICT LEVEL")
Prevalence_data <- inner_join(BCpop_district, BC_districtdata, by = c("SCHOOL_YEAR", "DISTRICT_NUMBER"), all.x = TRUE)

```

# Introduction
Educational policy can impact practice by reorganizing educational delivery and reconceptualizing the meaning and impact of education on society (Ball, 1998). Educational policy exists as a mechanism to drive changes in school buildings that result in socially important outcomes (e.g., increasing literacy acquisition through improved instruction, reducing instances of bullying through prevention programs). However, the relationship between policy and practice in Western formal systems of education is often unclear, misdirected, or atheoretical (Cohen et al., 2009). Leveraging the power of policy as a change agent for the delivery and conceptualization of education stands to benefit a significant proportion of students (Swanson, 2007).

When policy is enacted, the intended outcomes are often not realized for months, years, or decades (Gu et al., 2018). For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores (Luke & McArdle, 2009). Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest (Michelmore & Dynarski, 2017).

```{comprehensive display}

g_vic <- district_class %>%
  filter(DISTRICT_NAME == "Greater Victoria") %>%
  mutate(SCHOOL_YEAR = as.factor(SCHOOL_YEAR),
         SCHOOL_YEAR = as.numeric(SCHOOL_YEAR),
         SPECIAL_NEEDS_CODE = as.factor(SPECIAL_NEEDS_CODE),
          SPECIAL_NEEDS_CODE = as.numeric(SPECIAL_NEEDS_CODE),
         NUMBER_OF_STUDENTS = as.factor(NUMBER_OF_STUDENTS),
         NUMBER_OF_STUDENTS = as.numeric(NUMBER_OF_STUDENTS)) %>%
  select("SCHOOL_YEAR",
         "SPECIAL_NEEDS_CODE",
         "NUMBER_OF_STUDENTS") %>%
  as_tibble()

install.packages("GGally")
library(GGally)

#best one yet! Box-and-whisker plot
ggpairs(g_vic,
        upper = list(continuous = ggally_dot_and_box, combo = ggally_box_no_facet),
        lower = list (continuous = "smooth",
                      combo = "facetdensity",
                      mapping = aes(color = SPECIAL_NEEDS_CODE)))

#let's try bar graph
ggpairs(g_vic,
        upper = list(continuous = ggally_barDiag, combo = ggally_barDiag),
        lower = list (continuous = "smooth",
                      combo = "facetdensity",
                      mapping = aes(color = SPECIAL_NEEDS_CODE)))

ggally_dot_and_box(g_vic, 
                   mapping = ggplot2::aes(x = SCHOOL_YEAR, y = NUMBER_OF_STUDENTS),
                   boxPlot = TRUE) +
  theme(axis.text.x = element_blank())

```

## Current Study
Our project primarily examines two datasets detailing head counts of students with exceptionalities eligible for special education services aged 6-21. The datasets offer insights to the categorization for special education eligibility in public schools within British Columbia (BC) and Oregon (OR). The head counts from BC are collected from 2002/2003 to the most recent data from 2019/2020. The OR head counts are drawn from an annual dataset spanning 2002 through 2020. Levels of categorization include school-, district-, and provincial-level head counts for BC. The OR data set includes state-wide head counts that are not aggregated by school or district.

@@

Second, we hope to closely analyze any changes, or lack thereof, within the BC data following Canadian Supreme Court rulings between the British Columbia Teachers Federation and the province of BC (See Lambert, 2017). Canadian supreme court rulings have significant implications on educational policy in the provinces and, in 2016, resulted in a limit to the number of students with disability designations in classrooms resulting in an expansion of the number of specialist teachers’ that school districts are required to hire. The driving theory of change is that decreasing student-to-teacher ratios and capping the number of students with exceptionalities in each classroom causes student designation practices to change. Theoretically, school district administrators made decisions based on current staffing levels, student designations, diagnostic capacity and malleability, as well as external pressures to adjust practice according to legal mandates.

Finally, we will explore differences between urban and rural school districts in BC. Districts are categorized by population size and proximity to metropolitan areas, as recorded and defined by the Statistics Canada census from 2016. School district level data is useful for examining longitudinal trends for individual administrative units (e.g., superintendent’s office and director(s) of inclusive education); however, this level of data analysis cannot speak to province-wide trends in the culture of superordinate legal directives which impact swift policy changes. Given this, we will run analyses on both exemplar districts to highlight localized policy impacts on practice and amalgamated districts (e.g., school districts defined as rural by Statistics Canada) to investigate cultural shifts in the policy-to-practice association.

## Problem Statement and RQs
In comparison to the United States, Canadian education policy receives little attention and scholarly interest (Walker & Bergmann, 2013). While Canadian K-12 academic achievement outcomes are generally higher than a national-level comparison with the US, there are ongoing policy issues to address when provincial ministries of education are crafting legislation and procedures to impact a top-tier system of education. Significant events, such as the 2016 Canadian supreme court ruling that directly impacted British Columbia classrooms, have downstream effects on instructional interactions; yet a retroactive policy lens is rarely applied after such events occur. The current study seeks to address the knowledge gap of downstream instructional interact effects after a significant event in BC educational policy.

@@

# Methods

@@

The three datasets were used for analysis in R (cite R) for reproducibility and data visualization. All code for analysis and data representation is publicly available for reproducibility (https://github.com/JoeSwinehart/datascienceproject). There are several caveats for data use and cleaning in the current analysis. First, the BC and OR datasets were not a completely parallel comparison, because the OR data set includes a broader age range of students (ages 3 - 21) as compared to the BC data set (ages 5 - 19). The Oregon data set was pared down to include a similar age range to BC data; however, it is worth noting that differences in age range may represent different approaches to diagnostic categorization (e.g., intellectual disability in BC can be diagnosed at age 5; whereas the Oregon data follows IDEA where it is more likely young children will be categorized as developmentally delayed). Second, data masking necessitates a decision-point for analysis. Specifically, the BC Ministry of Education data masks all cells with headcounts below 10 students, therefore our analyses left out these cases when computing effect sizes for provincial-level data. Data masking was most prominently an issue when examining school district-level data in BC, as smaller school districts often masked low-incidence exceptionality data. When appropriate, masked data is assigned a conservative estimate of (n = 9) such that enough data units were included in effect size comparisons. Third, Canadian census data from 2016 is the basis of classification for district geographical location (e.g., urban, suburban, rural, exurban). Due to the geographical distribution of population in BC, only two categories are used for the current analyses: urban and rural. Fourth, due to low population numbers with independent schools in BC, the current study will only examine public school data for the years 2002 - 2019.

## Analysis Plan
The majority of the analyses for the current study are descriptive in nature. The descriptive analysis includes a table or visual representation of prevalence rates or headcounts of students with exceptionalities over time. As the current study is examining anomalies in trend data, trends that break the longitudinal trendline (e.g., sudden drop in prevalence after a decade of linear increase in prevalence) may be examined further. That is, effect sizes between two groups (e.g., group one is three years before a policy change, and group two is three years after a policy change) may be examined to detect the magnitude of differences between groups if they are hypothesized to represent a change in practice stemming from policy implementation. More specifically on the calculation of effect sizes, the Hedge’s g statistic will be used because of the robustness against unequal group sizes.

$$
Hedges' g = \frac{M_1 + M_2}{SD_{pooled}} 
$$
# Results
Writing on results here...

First, we see that the BC trends do not show significant changes after the 2016 supreme court ruling.
```{r Provinical_Data_Overview}

BCprovincial <- BCdatastripped %>% 
  filter(DATA_LEVEL == "PROVINCE LEVEL",
         PUBLIC_OR_INDEPENDENT == "PROVINCE - Total")  %>% 
  mutate(Public_population = BCpopulation$TOTAL_ENROLMENT[SCHOOL_YEAR-8],
         prevalence = NUMBER_OF_STUDENTS/Public_population) %>% 
  mutate(SCHOOL_YEAR = SCHOOL_YEAR+1995)

 ggplot(BCprovincial, aes(x = SCHOOL_YEAR, y = prevalence)) +
  geom_line(color = "cornflowerblue") +
   facet_wrap(~SPECIAL_NEEDS_CODE) +
   labs(title = "BC Provincial Prevalence of Students with Exceptionalities by Category",
        x = "School Year",
        y = "Prevalence Rate")
```

```{r cleaning Oregon data}
primodata <- read_excel(here("data", "primo_merged_oregon_data.xlsx"))
cleanest_primo <-primodata %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names()

cleanest_primo <- cleanest_primo %>% 
  select(-six_to_twentyone_total, -x24)

cleanest_primo2 <- cleanest_primo [!(cleanest_primo$disability=="TOTAL (SUM OF ALL OF THE ABOVE)" |cleanest_primo$disability=="OTHER HEALTH IMPAIRMENT" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY" | cleanest_primo$disability=="SPECIFIC LEARNING DISABILITY"),]
cleanest_primo2

fresh_primo <- cleanest_primo2 %>% 
  pivot_longer(starts_with("x"), names_to = "age", values_to = "frequency") %>% 
  mutate(age = parse_number(age), 
         year = parse_number(year))

fresh_primo %>% 
  group_by(year, disability) %>% 
  summarize(total=sum(frequency, na.rm=TRUE))

```

This is roughly in line with Oregon's designation data, showing that there were no state-wide anomalies after the year 2016.
```{r}
#this is more in line with the BC graphing
ggplot(fresh_primo, aes(year, frequency))+
 geom_smooth(color = "cornflowerblue") +
 theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
 facet_wrap(~disability)+
    labs(color="frequency", x= "Year", y="Student Count", title = "Oregon Headcount of Students with Exceptionalities by Category")

```

Even for diagnostically malleable categories, such as students classified in Oregon has having an "emotional disturbance," or in BC students requiring a "high behavioural need", Oregon and BC had highly similar trends.

``` {r BCOR combo}

combined <- fresh_primo %>% 
  as_tibble() %>% 
    mutate(ORheadcount = fct_recode(disability,
                          "Deafblind" = "DEAF-BLINDNESS",
                          "Moderate+ ID" = "INTELLECTUAL DISABILITY",
                          "High Behavioural Need" = "EMOTIONAL DISTURBANCE",
                          "Visual Impairment" = "VISUAL IMPAIRMENT",
                          "Deaf or Hard of Hearing" = "HEARING IMPAIRMENT",
                          "Autism" = "AUTISM",
                          "Mild ID" = "INTELLECTUAL DISABILITY")) %>% 
  filter(ORheadcount == "High Behavioural Need") %>% 
  group_by(year) %>% 
  summarise(frequency = sum(frequency))

BCAutism <- BCprovincial %>% 
  filter(SPECIAL_NEEDS_CODE == "High Behavioural Need")

ggplot() +
  geom_smooth(data = combined, aes(x = year, y = frequency, color = "Oregon")) +
  geom_smooth(data = BCAutism, aes(x = SCHOOL_YEAR, y = NUMBER_OF_STUDENTS, color = "British Columbia"), se = FALSE) +
  labs(title = "BC and Oregon Headcount of Students with Chronic Health Conditions",
        x = "School Year",
        y = "Headcount",
       color = "Legend",
        scale_color_manual(labels = c("Oregon", "British Columbia"), values = c("blue", "red")))

```

Then, we examine the rural versus urban trends and notice a few things...

```{r urban vs. rural}

bc_data_district_classification <- read_excel(here("data", "bc_data_district_classification.xlsx"))

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Exploring Anomalies in Pacific Northwest Special Education Trends},
  pdfauthor={David Fainstein1, Janette Avelar1, Makayla Whitney1, \& Joe Swinehart1},
  pdflang={en-EN},
  pdfkeywords={Vancouver, British Columbia, Oregon, Special Education, SpEd},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{PACNW SpEd Trends}
\keywords{Vancouver, British Columbia, Oregon, Special Education, SpEd\newline\indent Word count: 0<n<1,000,000}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

\title{Exploring Anomalies in Pacific Northwest Special Education Trends}
\author{David Fainstein\textsuperscript{1}, Janette Avelar\textsuperscript{1}, Makayla Whitney\textsuperscript{1}, \& Joe Swinehart\textsuperscript{1}}
\date{}

\authornote{

Janette Avelar, University of Oregon College of Education

@@

Joe Swinehart, Behavioral Research \& Teaching

@@

Contact: \href{mailto:dfainste@uoregon.edu}{\nolinkurl{dfainste@uoregon.edu}}

The authors made the following contributions. David Fainstein: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Janette Avelar: Writing - Review \& Editing; Makayla Whitney: Writing - Review \& Editing; Joe Swinehart: Writing - Review \& Editing.

Correspondence concerning this article should be addressed to David Fainstein, 1655 Alder, Eugene, Oregon, USA 97403. E-mail: \href{mailto:dfainste@uoregon.edu}{\nolinkurl{dfainste@uoregon.edu}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} University of Oregon}

\abstract{
This paper will examine the changes in public school special education populations in Vancouver, British Columbia and Oregon over the last 20 years. The British Columbia (BC) data was pulled from an open-source data catalogue, published by the education analytics division of the Ministry of Education in British Columbia. This longitudinal data acquistion is part of formal student status reporting that is required within BC public schools - typically occuring in October of each school year. The Oregon data was pulled from the Oregon Department of Education and covers special education enrollment from 2002 - 2020. This data acquistion is part of formal student status reporting that is required for compliance with federal IDEA laws - typically occuring in November or February of each school year. Within the BC dataset, students with exceptionalities are categorized in one of 12 eligibility categories; within the Oregon dataset, this is 13 categories.
}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Educational policy can impact practice by reorganizing educational delivery and reconceptualizing the meaning and impact of education on society (Ball, 1998). Educational policy exists as a mechanism to drive changes in school buildings that result in socially important outcomes (e.g., increasing literacy acquisition through improved instruction, reducing instances of bullying through prevention programs). However, the relationship between policy and practice in Western formal systems of education is often unclear, misdirected, or atheoretical (Cohen et al., 2009). Leveraging the power of policy as a change agent for the delivery and conceptualization of education stands to benefit a significant proportion of students (Swanson, 2007).

When policy is enacted, the intended outcomes are often not realized for months, years, or decades (Gu et al., 2018). For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores (Luke \& McArdle, 2009). Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest (Michelmore \& Dynarski, 2017).

\hypertarget{current-study}{%
\subsection{Current Study}\label{current-study}}

Our project primarily examines two datasets detailing head counts of students with exceptionalities eligible for special education services aged 6-21. The datasets offer insights to the categorization for special education eligibility in public schools within British Columbia (BC) and Oregon (OR). The head counts from BC are collected from 2002/2003 to the most recent data from 2019/2020. The OR head counts are drawn from an annual dataset spanning 2002 through 2020. Levels of categorization include school-, district-, and provincial-level head counts for BC. The OR data set includes state-wide head counts that are not aggregated by school or district.

@@

Second, we hope to closely analyze any changes, or lack thereof, within the BC data following Canadian Supreme Court rulings between the British Columbia Teachers Federation and the province of BC (See Lambert, 2017). Canadian supreme court rulings have significant implications on educational policy in the provinces and, in 2016, resulted in a limit to the number of students with disability designations in classrooms resulting in an expansion of the number of specialist teachers' that school districts are required to hire. The driving theory of change is that decreasing student-to-teacher ratios and capping the number of students with exceptionalities in each classroom causes student designation practices to change. Theoretically, school district administrators made decisions based on current staffing levels, student designations, diagnostic capacity and malleability, as well as external pressures to adjust practice according to legal mandates.

Finally, we will explore differences between urban and rural school districts in BC. Districts are categorized by population size and proximity to metropolitan areas, as recorded and defined by the Statistics Canada census from 2016. School district level data is useful for examining longitudinal trends for individual administrative units (e.g., superintendent's office and director(s) of inclusive education); however, this level of data analysis cannot speak to province-wide trends in the culture of superordinate legal directives which impact swift policy changes. Given this, we will run analyses on both exemplar districts to highlight localized policy impacts on practice and amalgamated districts (e.g., school districts defined as rural by Statistics Canada) to investigate cultural shifts in the policy-to-practice association.

\hypertarget{problem-statement-and-rqs}{%
\subsection{Problem Statement and RQs}\label{problem-statement-and-rqs}}

In comparison to the United States, Canadian education policy receives little attention and scholarly interest (Walker \& Bergmann, 2013). While Canadian K-12 academic achievement outcomes are generally higher than a national-level comparison with the US, there are ongoing policy issues to address when provincial ministries of education are crafting legislation and procedures to impact a top-tier system of education. Significant events, such as the 2016 Canadian supreme court ruling that directly impacted British Columbia classrooms, have downstream effects on instructional interactions; yet a retroactive policy lens is rarely applied after such events occur. The current study seeks to address the knowledge gap of downstream instructional interact effects after a significant event in BC educational policy.

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

The three datasets were used for analysis in R (cite R) for reproducibility and data visualization. All code for analysis and data representation is publicly available for reproducibility (\url{https://github.com/JoeSwinehart/datascienceproject}). There are several caveats for data use and cleaning in the current analysis. First, the BC and OR datasets were not a completely parallel comparison, because the OR data set includes a broader age range of students (ages 3 - 21) as compared to the BC data set (ages 5 - 19). The Oregon data set was pared down to include a similar age range to BC data; however, it is worth noting that differences in age range may represent different approaches to diagnostic categorization (e.g., intellectual disability in BC can be diagnosed at age 5; whereas the Oregon data follows IDEA where it is more likely young children will be categorized as developmentally delayed). Second, data masking necessitates a decision-point for analysis. Specifically, the BC Ministry of Education data masks all cells with headcounts below 10 students, therefore our analyses left out these cases when computing effect sizes for provincial-level data. Data masking was most prominently an issue when examining school district-level data in BC, as smaller school districts often masked low-incidence exceptionality data. When appropriate, masked data is assigned a conservative estimate of (n = 9) such that enough data units were included in effect size comparisons. Third, Canadian census data from 2016 is the basis of classification for district geographical location (e.g., urban, suburban, rural, exurban). Due to the geographical distribution of population in BC, only two categories are used for the current analyses: urban and rural. Fourth, due to low population numbers with independent schools in BC, the current study will only examine public school data for the years 2002 - 2019.

\hypertarget{analysis-plan}{%
\subsection{Analysis Plan}\label{analysis-plan}}

The majority of the analyses for the current study are descriptive in nature. The descriptive analysis includes a table or visual representation of prevalence rates or headcounts of students with exceptionalities over time. As the current study is examining anomalies in trend data, trends that break the longitudinal trendline (e.g., sudden drop in prevalence after a decade of linear increase in prevalence) may be examined further. That is, effect sizes between two groups (e.g., group one is three years before a policy change, and group two is three years after a policy change) may be examined to detect the magnitude of differences between groups if they are hypothesized to represent a change in practice stemming from policy implementation. More specifically on the calculation of effect sizes, the Hedge's g statistic will be used because of the robustness against unequal group sizes.

\[
Hedges' g = \frac{M_1 + M_2}{SD_{pooled}} 
\]
\# Results
Writing on results here\ldots{}

First, we see that the BC trends do not show significant changes after the 2016 supreme court ruling.
\includegraphics{Final_project_files/figure-latex/Provinical_Data_Overview-1.pdf}

\begin{verbatim}
## # A tibble: 198 x 18
##    year  disability    x6    x7    x8    x9   x10   x11   x12   x13   x14   x15
##    <chr> <chr>      <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
##  1 2002~ INTELLECT~   135   172   220   261   279   303   370   364   388   378
##  2 2002~ HEARING I~    46    66    69    67    79    78    69    70    53    72
##  3 2002~ SPEECH OR~  2502  2466  2443  2205  1894  1282   874   677   493   340
##  4 2002~ VISUAL IM~    18    27    25    16    12    26    19    24    31    19
##  5 2002~ EMOTIONAL~    81   178   241   292   347   433   504   487   549   548
##  6 2002~ ORTHOPEDI~    65    77    64    49    64    71    50    48    44    43
##  7 2002~ DEAF-BLIN~     0     2     0     0     1     0     2     0     4     2
##  8 2002~ MULTIPLE ~     0     0     0     0     0     0     0     0     0     0
##  9 2002~ AUTISM       267   289   290   334   325   316   290   256   243   219
## 10 2002~ TRAUMATIC~     6    11    17    23    22    19    21    23    36    36
## # ... with 188 more rows, and 6 more variables: x16 <dbl>, x17 <dbl>,
## #   x18 <dbl>, x19 <dbl>, x20 <dbl>, x21 <dbl>
\end{verbatim}

\begin{verbatim}
## # A tibble: 198 x 3
## # Groups:   year [18]
##     year disability                    total
##    <dbl> <chr>                         <dbl>
##  1  2002 AUTISM                         3339
##  2  2002 DEAF-BLINDNESS                   17
##  3  2002 DEVELOPMENTAL DELAY               0
##  4  2002 EMOTIONAL DISTURBANCE          4736
##  5  2002 HEARING IMPAIRMENT              873
##  6  2002 INTELLECTUAL DISABILITY        4387
##  7  2002 MULTIPLE DISABILITIES             0
##  8  2002 ORTHOPEDIC IMPAIRMENT           754
##  9  2002 SPEECH OR LANGUAGE IMPAIRMENT 15784
## 10  2002 TRAUMATIC BRAIN INJURY          306
## # ... with 188 more rows
\end{verbatim}

This is roughly in line with Oregon's designation data, showing that there were no state-wide anomalies after the year 2016.

\begin{verbatim}
## Warning: Removed 190 rows containing non-finite values (stat_smooth).
\end{verbatim}

\includegraphics{Final_project_files/figure-latex/unnamed-chunk-1-1.pdf}

Even for diagnostically malleable categories, such as students classified in Oregon has having an \enquote{emotional disturbance,} or in BC students requiring a \enquote{high behavioural need}, Oregon and BC had highly similar trends.

\includegraphics{Final_project_files/figure-latex/BCOR combo-1.pdf}

Then, we examine the rural versus urban trends and notice a few things\ldots{}

\begin{verbatim}
## Warning in FUN(newX[, i], ...): NAs introduced by coercion
\end{verbatim}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:District Census Population Table}District Classification by Population According to 2016 Census}
